---
title: "FitnessTracker"
output: html_document
date: "2024-01-21"
---

```{r}
library(tidyverse)
library(here)
library(skimr)
library(janitor)
library(lubridate)
library(ggplot2)

```
#read the files

```{r}
daily_activity <- read_csv("dailyActivity.csv")
View(daily_activity)

sleepDay <- read_csv("sleepDay.csv")
View(sleepDay)

hourlyIntensities<-read_csv(hourlyIntensities.csv)
View(hourlyIntensities)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

#Fixing formatting
```{r}
#SleepDay
sleepDay$SleepDay = as.POSIXct(sleepDay$SleepDay, format = "%m/%d/%Y %H:%M:%S", tz = Sys.timezone())
sleepDay$date<-format(sleepDay$SleepDay,format="%m/%d/%Y")
sleepDay$time<-format(sleepDay$SleepDay,format="%H:%M:%S")
```


```{r}
#hourlyintensities
hourlyIntensities$ActivityHour=as.POSIXct(hourlyIntensities$ActivityHour, format="%m/%d/%Y %I:%M:%S %p", tz=Sys.timezone())
hourlyIntensities$time <- format(hourlyIntensities$ActivityHour, format = "%H:%M:%S")
hourlyIntensities$date <- format(hourlyIntensities$ActivityHour, format = "%m/%d/%y")

```

##Preview the data

```{r}
#dailyActivity
head(daily_activity)
str(daily_activity)
```
```{r}
head(hourlyIntensities)
```




```{r}
#SleepDay
str(sleepDay)
```

#finding distinct users in each data frame
```{r}
n_distinct(daily_activity$Id)
n_distinct(sleepDay$Id)

```

#summary statistics
```{r}
daily_activity %>%
  select(TotalDistance,
         VeryActiveDistance,
         SedentaryActiveDistance,
         LightActiveDistance,
         VeryActiveMinutes,
         SedentaryMinutes,
         LightlyActiveMinutes) %>%
  summary()

sleepDay %>%
  select(TotalMinutesAsleep,
         TotalTimeInBed) %>%
  summary()


```
#visualizing the data

```{r}
mean_values <- colMeans(daily_activity[, c("VeryActiveMinutes", "FairlyActiveMinutes", "LightlyActiveMinutes", "SedentaryMinutes")])
#Create a data frame for ggplot
activity_data <- data.frame(
  Activity = c("Very Active", "Fairly Active", "Lightly Active", "Sedentary"),
  MeanMinutes = mean_values)

# Plot using ggplot with rotated X-axis labels
ggplot(activity_data, aes(x = reorder(Activity, MeanMinutes), y = MeanMinutes, fill = Activity)) +
  geom_col(color = "black") +
  labs(title = "Mean Activity Minutes", x = "Activity Type", y = "Average Minutes") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


```{r}
avg_calories <- daily_activity %>%
  group_by(Id) %>%
  summarise(avg_Calories = mean(Calories))

# Categorize participants based on average calories
avg_calories <- mutate(avg_calories, WeightCategory = case_when(
  avg_Calories >= 1600 & avg_Calories <= 2200 ~ 'Maintain Weight',
  avg_Calories < 1600 ~ 'Gain Weight',
  avg_Calories > 2200 ~ 'Lose Weight'
))

# Plot the bar chart
ggplot(avg_calories, aes(x = "", y = avg_Calories, fill = WeightCategory)) +
  geom_bar(stat = 'identity', width = 1, color = 'white') +
  coord_polar("y") +
  labs(title = 'Average Calories and Weight Category by Participant', fill = 'Weight Category') +
  theme_void() 

```
#sleepday


```{r}
ggplot(sleepDay, aes(x = TotalMinutesAsleep, y = TotalTimeInBed)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Total Minutes Asleep vs. Total Time in Bed")

```
it's intriguing to note the concentration of data points where individuals spent more time in bed than the actual duration of sleep. 
we can remind individuals to limit screen time before bedtime. 
Suggest establishing a calming bedtime routine to help individuals unwind before sleep. This may include activities such as reading, gentle stretching, or practicing relaxation techniques.

```{r}
sleepDay <- mutate(sleepDay, SleepCategory = case_when(TotalMinutesAsleep >= 420 & TotalMinutesAsleep <= 540 ~ 'Recommended Sleep',TotalMinutesAsleep >= 360 & TotalMinutesAsleep < 420 | TotalMinutesAsleep > 540 ~ 'Appropriate Sleep',TotalMinutesAsleep < 360 ~ 'Inadequate Sleep',
  TRUE ~ 'Other' ))

ggplot(sleepDay, aes(x = SleepCategory, fill = SleepCategory)) + geom_bar() +labs(title = 'Sleep Categories', x = 'Sleep Category', y = 'Count') + theme_minimal()
```

```{r}
colnames(daily_activity)
```
#Merging the two datasets together

```{r}

activity_sleep_combined <- merge(daily_activity, sleepDay, by = "Id", all = TRUE)
```

```{r}
#View(activity_sleep_combined)
n_distinct(activity_sleep_combined$Id)
```
#Total Minutes Asleep vs. Total Steps
```{r}
ggplot(data = activity_sleep_combined, aes(x = SleepDay, y = TotalSteps, fill = SleepDay)) + 
  geom_bar(stat = "summary", fun = "mean") +
  xlab("Sleep Day") +
  ylab("Mean Total Steps") +
  ggtitle("Mean Total Steps by Sleep Day")

```
there really is no relationship between sleeps and steps per day.

```{r}
ggplot(data=activity_sleep_combined, aes(x=TotalSteps, y=Calories)) + 
  geom_point() + 
  geom_smooth() + 
  labs(title="Total Steps vs. Calories")
```
the more steps a person takes, the more calories they burn. It's a great idea to remind users of that to reach their presumed goals.

```{r}
ggplot(activity_data, aes(x = ActivityType, y = Minutes, fill = ActivityType)) +
  geom_violin() +
  xlab("Activity Type") +
  ylab("Minutes") +
  ggtitle("Total Minutes of Activity by Type") +
  scale_fill_manual(values = c("red", "yellow", "brown", "green")) +
  theme_minimal() +
  coord_flip() 
```
This visualization indicates that there is significant potential for promoting and encouraging physical activity.
```{r}
int_new <- hourlyIntensities %>%
  group_by(time) %>%
  drop_na() %>%
  summarise(mean_total_int = mean(TotalIntensity))

ggplot(data=int_new, aes(x=time, y=mean_total_int)) + geom_histogram(stat = "identity", fill='violet') +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title="Average Total Intensity vs. Time")
```
It's evident that individuals are most active between 5 pm and 7 pm, with additional peaks observed around 12 pm to 2 pm. These time intervals align with typical lunch breaks and the conclusion of the workday during the week. These periods present opportune moments to promote and encourage activities such as workouts, jogging, or taking a pet for a walk.

```{r}

```

```{r}

```